Nested Stochastic Block Models Applied to the Analysis of Single Cell Data

### 2021-04-15 manuscript.tex
updated introduction and random data

\begin{document}
\title{Stochastic Block Models Applied to the Analysis of Single Cell Data}
\author[1,2]{Leonardo Morelli}
\author[1,3]{Valentina Giansanti}
\author[1]{Davide Cittaro}
\affil[1]{Center for Omics Sciences, IRCCS San Raffaele Institute, Milan, Italy}
\affil[2]{Universit\`{a} Vita-Salute San Raffaele, Milan, Italy} 
\affil[3]{Department of Informatics, Systems and Communication, University of Milano-Bicocca, Milan, Italy}
\maketitle

@@

A key step in the analysis of scRNA-seq data and, more in general, of single cell data, is the identification of cell populations, groups of cells sharing similar properties. Several approaches have been proposed to achieve this task, based on well established clustering techniques \cite{wang_2017, lin_2017}, consensus clustering \cite{huh_2020, kiselev_2017} and deep learning \cite{li_2020}; many more have been recently reviewed \cite{krzak_2019, kiselev_2019} and benchmarked \cite{du_2018}. As the popularity of single cell analysis frameworks Seurat \cite{butler_2018} and Scanpy \cite{wolf_2018} raised, methods based instead on graph partitioning became the \emph{de facto} standards. Such methods require the construction of a cell neighbourhood graph (\emph{e.g.} by \emph{k }Nearest Neighbours, \emph{k}NN) which is then partitioned into communities; the latter step is typically performed using the Louvain method \cite{blondel_2008}, a fast algorithm for optimisation of graph modularity. While fast, this method does not guarantee that small communities in large networks are well defined. To overcome its limits, a more recent approach, the Leiden algorithm \cite{traag_2019}, has been implemented and it has been quickly adopted in the analysis of single cell data, for example by Scanpy and PhenoGraph \cite{levine_2015}. In addition to Newman's modularity \cite{newman_2004}, other definitions currently used in single cell analysis make use of a resolution parameter \cite{traag_2011, reichardt_2006} . In lay terms, resolution works as a threshold on the density within communities: lowering the resolution results in less and sparser communities and \emph{viceversa}. Identification of an appropriate resolution has been recognised as a major issue \cite{lhnemann_2020}, also because it requires the definition of a mathematical property (clusters) over biological entities (the cell groups), with little formal description of the latter. In addition, the larger the dataset, the harder is to identify small cell groups, as a consequence of the well-known resolution limit \cite{fortunato_2007}. Moreover, it has been demonstrated that random networks can have modularity \cite{guimer_2004} and its optimisation is incapable of separating actual structure from those arising simply of statistical fluctuations of the null model. Additional solutions to cell group identification from neighbourhood graphs have been proposed, introducing resampling techniques \cite{baran_2019} or clique analysis \cite{xu_2015}. Lastly, it has been proposed that high resolution clustering, \emph{e.g.} obtained with Leiden or Louvain methods, can be refined in agglomerative way using machine learning techniques \cite{miao_2020}.

@@

\begin{MPEquation}[!ht]
\begin{equation}
\Sigma = -\ln P(\boldsymbol A|\boldsymbol\theta, \boldsymbol b) - \ln P(\boldsymbol\theta, \boldsymbol b)
\end{equation}
\label{MPEquationElement:676396C0-2D6F-4084-89B1-951DE3032087}
\end{MPEquation}
so that inference is performed by minimizing the information required to describe the data (Occam's razor); \emph{graph-tool} is able to efficiently do this by a Markov Chain Monte Carlo approach \cite{peixoto_2014}. SBM itself may fail to identify small groups in large graphs, hence hierarchical formulation has been proposed \cite{peixoto_2014_h}. Under this model, communities are agglomerated at a higher level in a block multigraph, also modelled using SBM. This process is repeated recursively until a graph with a single block is reached, creating a Nested Stochastic Block Model (nSBM).

In this work we propose nSBM for the analysis of single cell data, in particular scRNA-seq data. This approach identifies cell groups in a statistical robust way and, moreover, it is able to determine the likelihood of the grouping, thus allowing model selection. In addition, it is possible to measure the confidence of assignment to groups. We show that such information may be exploited to perfect the notion of cell groups and the identification of markers.

@@

\emph{schist} is a convenient wrapper to the \emph{graph-tool} python library, written in python and designed to be used with \emph{scanpy}. The most prominent function is \emph{schist.inference.nested\_model()} which takes a \emph{AnnData} object as input and fits a nested stochastic block model on the \emph{k}NN graph built with \emph{scanpy} functions (\emph{e.g.\ scanpy.tools.neighbors()}). When launched with default parameters, \emph{schist} fits a model which maximises the posterior probability of having a set of cell groups (or blocks) given a graph. \emph{schist} annotates cells in the data object with all the groups found at each level of a hierarchy. As there could be more model fits with similar entropy, \emph{schist} could explore the space of solutions with a Markov Chain Monte Carlo (MCMC) algorithm, to perform model averaging; this step is performed until it converges, that is the difference in model entropy in \emph{n} continuous iterations remains under a specified threshold. Sampling from the posterior distribution can be used to study the distribution of the number of groups, at each level. This information (group marginals) could be studied to identify the most probable number of cell groups.
Once \emph{schist} has fitted a model, it evaluates the difference in entropy given by assigning every cell to every possible group. This step generates the matrix of \emph{cell affinity}, that is the probability for a cell to belong to a specific group. A cell affinity matrix is generated and returned for every hierarchy level. We show that this information can be useful to evaluate cluster consistence.

%As the core object of \emph{graph-tool }used by \emph{schist}, the \emph{NestedBlockState}, cannot be serialised by \emph{scanpy}'s IO infrastructure, we implemented simple read\slash write functions to dump \emph{AnnData} is h5ad format, together with the \emph{NestedBlockState} as a separate file using python's \emph{pickle} library.

@@

\textcolor{red}{TO BE UPDATED}

One of the most relevant difference between nSBM and other methods to cluster single cells is that it relies on robust statistical modelling. In this sense, the number of groups identified strictly mirrors the amount of information contained in the data. An important consequence is that absence of information can be properly handled. To show this property we performed a simple experiment on a randomised \emph{k}NN graph. To this end we collected data for 3k PBMC (available as preprocessed data in \emph{scanpy}) and shuffled the edges of the prebuilt \emph{k}NN graph, this to keep the general graph properties unchanged. We tested that the degree distribution does not change after randomisation (Kolmogorov-Smirnov $D=0.0431$, $p=0.993$). We found that the default strategy, based on maximisation of modularity, identifies 24 cell groups, whereas nSBM does not identify any cell group.

\begin{figure}[H]
\centering
\includegraphics[keepaspectratio,width=\textwidth,height=0.4\textheight]{FIgure_Random.png}
\caption[]{nSBM results for randomised data. (A) UMAP embedding of ~3k PBMC after standard processing with Leiden approach (left) or nSBM (right). Cell grouping is consistent for both the approaches ($ARI=0.804$). (B) UMAP embedding of the same data after randomisation of \emph{k}NN graph edges. In this case nSBM does not return any cell grouping, while optimisation of modularity finds up to 24 different cell groups}\label{FigureRandom}
\end{figure}

>>>>

\begin{document}
\title{Nested Stochastic Block Models Applied to the Analysis of Single Cell Data}
\author[1,2]{Leonardo Morelli}
\author[1,3]{Valentina Giansanti}
\author[1]{Davide Cittaro}
\affil[1]{Center for Omics Sciences, IRCCS San Raffaele Institute, Milan, Italy}
\affil[2]{Universit\`{a} Vita-Salute San Raffaele, Milan, Italy} 
\affil[3]{Department of Informatics, Systems and Communication, University of Milano-Bicocca, Milan, Italy}
\maketitle

@@

A key step in the analysis of scRNA-seq data and, more in general, of single cell data, is the identification of cell populations, that is groups of cells sharing similar properties. Several approaches have been proposed to achieve this task, based on well established clustering techniques \cite{wang_2017, lin_2017}, consensus clustering \cite{huh_2020, kiselev_2017, Ranjan_2021} and deep learning \cite{li_2020}; many more have been recently reviewed \cite{krzak_2019, kiselev_2019} and benchmarked \cite{du_2018}. As the popularity of single cell analysis frameworks Seurat \cite{butler_2018} and Scanpy \cite{wolf_2018} raised, methods based instead on graph partitioning became the \emph{de facto} standards. Such methods require the construction of a cell neighbourhood graph (\emph{e.g.} by \emph{k }Nearest Neighbours, \emph{k}NN or shared Nearest Neighbors, \emph{s}NN). Encoding cell-to-cell similarities into graphs has practical advantages, as many algorithms for graph analysis can be applied and interpreted in a biological way. A notable example is the analysis of cell trajectories which can be derived from the analysis of random walks on the NN graph (\textcolor{red}{CHECK THIS: it's actually from diffusion maps}; computation of RNA moments in scRNA velocity is also based on the NN graph structure \cite{Bergen_2020}. Arguably, the biggest utility of NN structure is the possibility to identify cell groups by partitioning the graph into communities; the latter step is typically performed using the Louvain method \cite{blondel_2008}, a fast algorithm for optimisation of graph modularity. While fast, this method does not guarantee that small communities in large networks are well defined. To overcome its limits, a more recent approach, the Leiden algorithm \cite{traag_2019}, has been implemented and it has been quickly adopted in the analysis of single cell data, for example by Scanpy \cite{wolf_2018} and PhenoGraph \cite{levine_2015}. In addition to Newman's modularity \cite{newman_2004}, other definitions currently used in single cell analysis make use of a resolution parameter \cite{traag_2011, reichardt_2006}. In lay terms, resolution works as a threshold on the density within communities: lowering the resolution results in less and sparser communities and \emph{viceversa}. Identification of an appropriate resolution has been recognised as a major issue \cite{lhnemann_2020}, also because it requires the definition of a mathematical property (clusters) over biological entities (the cell groups), with little formal description of the latter. In addition, the larger the dataset, the harder is to identify small cell groups, as a consequence of the well-known resolution limit \cite{fortunato_2007}. Moreover, it has been demonstrated that random networks can have modularity \cite{guimer_2004} and its optimisation is incapable of separating actual structure from those arising simply of statistical fluctuations of the null model. Additional solutions to cell group identification from neighbourhood graphs have been proposed, introducing resampling techniques \cite{baran_2019, Tang_Sackton_2020} or clique analysis \cite{xu_2015}. Lastly, it has been proposed that high resolution clustering, \emph{e.g.} obtained with Leiden or Louvain methods, can be refined in agglomerative way using machine learning techniques \cite{miao_2020}.

@@

\begin{MPEquation}[!ht]
\begin{equation}
\Sigma = -\ln P(\boldsymbol A|\boldsymbol\theta, \boldsymbol b) - \ln P(\boldsymbol\theta, \boldsymbol b)
\end{equation}
\label{MPEquationElement:676396C0-2D6F-4084-89B1-951DE3032087}
\end{MPEquation}

so that inference is performed by minimizing the information required to describe the data (Occam's razor); \emph{graph-tool} is able to efficiently do this by a Markov Chain Monte Carlo approach \cite{peixoto_2014}. SBM itself may fail to identify small groups in large graphs, hence hierarchical formulation has been proposed \cite{peixoto_2014_h}. Under this model, communities are agglomerated at a higher level in a block multigraph, also modelled using SBM. This process is repeated recursively until a graph with a single block is reached, creating a Nested Stochastic Block Model (nSBM).

In this work we propose nSBM for the analysis of single cell data, in particular scRNA-seq data. This approach identifies cell groups in a statistical robust way and, moreover, it is able to determine the likelihood of the grouping, thus allowing model selection. In addition, it is possible to measure the confidence of assignment to groups. \textcolor{red}{We show that such information may be exploited to perfect the notion of cell groups and the identification of markers.}

@@

\emph{schist} is a convenient wrapper to the \emph{graph-tool} python library, written in python and designed to be used with \emph{scanpy}. The most prominent function is \emph{schist.inference.nested\_model()} which takes a \emph{AnnData} object as input and fits a nested stochastic block model on the \emph{k}NN graph built with \emph{scanpy} functions (\emph{e.g.\ scanpy.tools.neighbors()}). When launched with default parameters, \emph{schist} fits a model which maximises the posterior probability of having a set of cell groups (or blocks) given a graph. \emph{schist} then annotates cells in the data object with all the groups found at each level of a hierarchy. Given the large size of the NN graph, it is possible that a single solution represents local minima of the fitting process. In addition, it is possible that multiple solutions are equally acceptable to represent the graph partitioning and a better description is given by the consensus over such solutions \cite{Peixoto_2021_consensus}. To overcome these issues, \emph{schist} fits multiple instances in parallel and returns the inferred consensus model, alongside the marginal probabilities for each cell to belong to a specific group (\emph{cell marginals}). 
In addition to the default Nested Stochastic Block Model, \emph{schist} implements the possibility to fit a Bayesian model able to find statistically significant assortative modules \cite{Zhang_Peixoto_2020}, eliminating the need to set a resolution parameter as required in standard community detection by maximization of modularity.
Once a model has been fitted, \emph{schist} allows to evaluate the difference in entropy given by assigning every cell to every possible group. This step allows to estimate the probability for a cell to belong to a specific group (\emph{cell affinity}). We show that cell affinity may have practical application in common tasks in single cell analysis, such as label transfer.

@@

One of the most relevant difference between nSBM and other methods to cluster single cells is that it relies on robust statistical modelling. In this sense, the number of groups identified strictly mirrors the amount of information contained in the data. An important consequence is that absence of information can be properly handled. To show this property we performed a simple experiment on a randomised \emph{k}NN graph. To this end we collected data for 3k PBMC (available as preprocessed data in \emph{scanpy}) and shuffled the edges of the prebuilt \emph{k}NN graph, this to keep the general graph properties unchanged. We tested that the degree distribution does not change after randomisation (Kolmogorov-Smirnov $D=0.0733$, $p=0.703$). We found that the default strategy, based on maximisation of modularity, identifies 24 cell groups, whereas nSBM does not identify any cell group.

\begin{figure}[H]
\centering
\includegraphics[keepaspectratio,width=\textwidth,height=0.4\textheight]{FIgure_Random.png}
\caption[]{nSBM results for randomised data. (A) UMAP embedding of ~3k PBMC after standard processing with Leiden approach (left) or nSBM (right). Cell grouping is consistent for both the approaches (Adjusted Rand Index $ARI=0.869$). (B) UMAP embedding of the same data after randomisation of \emph{k}NN graph edges. In this case nSBM does not return any cell grouping, while optimisation of modularity finds up to 24 different cell groups}\label{FigureRandom}
\end{figure}
